The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-06 manuscript/manuscript.Rnw
fix TMSB4 error in dog
Fix elephant paragraph

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB10} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of the annexin complex were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_evolution_tree}. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the elephant are shown in Table 6 with a list of all genes in
Supplemental Table X. The gene with the highest expression
(\ac{FPKM}=33439.5) was the natriuretic peptide precursor B (NPPB). The
gene with the second highest expression (\ac{FPKM}=20233) was the microRNA
miR-675, which is imbedded in the first exon of the maternally
imprinted long non-coding RNA gene H19 (Keniry et al., 2012), a gene
also highly expressed in the elephant placenta (\ac{FPKM}=7164.3). Five
uncharacterized genes are expressed to a high degree in the elephant
placenta (Table 6), three of which are a member of an
elephant-specific expansion of 11 genes closely related to the
thymus-specific serine protease gene PRSS16. The other 8 PRSS16-like
genes are also expressed to a lesser degree (FPKM<284.014). Two other
novel genes with high expression are related to the gene SSLP1 (Table
6), which is also expressed in the cow and dog placentas (Tables 4 and
5). Placental expression of the hormones relaxin (RLN1) and prolactin
(PRL) are exceptionally high in the elephant when compared to other
mammals (Tables 6 and 8). There is little to no placental expression
of the estrogen synthase gene, CYP19A1, (FPKM<1) and the enzyme that
produces progesterone HSD3B1 (\ac{FPKM}=0). Other genes with high
expression in the elephant placenta include genes involved in
oxidative stress (TXN), growth (IGF2), and PAEP, a gene usually
expressed in the endometrium of the uterus in humans (Rutanen et al.,
1987).

>>>>

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB4} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of the annexin complex were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_evolution_tree}. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
with a list of all genes in
\cref{tab:all_loxodonta_africana_placenta}. The gene with the
highest expression was the natriuretic peptide precursor B
(\Sexpr{gene.expression("loxodonta africana","NPPB",complete=TRUE)}).
Two thymosins, %
\Sexpr{gene.expression("loxodonta africana","TMSB10",complete=TRUE)} %
and %
\Sexpr{gene.expression("loxodonta africana","TMSB4Y",complete=TRUE)}
were also also was highly expressed. The miRNA mir-675
(\Sexpr{gene.expression("loxodonta
  africana","ENSLAFG00000027406",complete=TRUE)}). which is imbedded
in the first exon of the maternally imprinted long non-coding RNA gene
\textit{H19} \cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene
which is currently un-annotated in \elephant{}. %
\DLA{Probably consider fixing the unannotation of H19; it seems pretty
  obvious to me, though it would require fixing the GTF file for
  elephant with the right 
  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan prediction} for this region}
Five uncharacterized genes are expressed to a high degree in the
\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
which %
%
(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000028493",complete=TRUE)},
\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000032011",complete=TRUE)},
\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000031933",complete=TRUE)}) %
%
are members of an \elephant{}-specific expansion of 11 genes closely
related to the thymus-specific serine protease gene \textit{PRSS16}
\cref{fig:prss16_evolution_tree} The other 8 \textit{PRSS16}-like
genes are also expressed to a lesser degree %
(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000026505",complete=TRUE)}). %
%
Two other novel genes with high expression are related to the gene
\textit{SSLP1} %
%
(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000032659",complete=TRUE)} %
and %
%
\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000029551",complete=TRUE)},
%
\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
Placental expression of the hormones relaxin (\textit{RLN}, %
\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000036090")})
and prolactin (\textit{PRL}) are exceptionally high in the elephant
when compared to other mammals
(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
is little to no placental expression of the estrogen synthase gene,
\textit{CYP19A1} %
(\Sexpr{gene.expression("loxodonta africana","CYP19A1")}) and the
enzyme that produces progesterone HSD3B1 %
(\Sexpr{gene.expression("loxodonta africana","HSD3B")}). Other genes
with high expression in the elephant placenta include genes involved
in oxidative stress %
(\Sexpr{gene.expression("loxodonta africana","TXN",complete=TRUE)}),
growth %
(\Sexpr{gene.expression("loxodonta africana","IGF2",complete=TRUE)})
and %
\Sexpr{gene.expression("loxodonta africana","PAEP",complete=TRUE)}, %
a gene usually expressed in the endometrium of the uterus in humans
\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.

<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of \textit{RLN2} and related genes}
\label{fig:rln2_evolution_tree}
\end{figure}
